---
title: "Multiple linear regression"
author: ""
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

#### Title: CUNY SPS MDS DATA606_LAB9"
#### Author: Charles Ugiagbe
#### Date: 11/6/2021

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate the course and the instructor anonymously. However, the use of these student evaluations as an indicator of course quality and teaching effectiveness is often criticized because these measures may reflect the influence of non-teaching related characteristics, such as the physical appearance of the instructor. The article titled, "Beauty in the classroom: instructors' pulchritude and putative pedagogical productivity" by Hamermesh and Parker found that instructors who are viewed to be better looking receive higher instructional ratings. 

Here, you will analyze the data from this study in order to learn what goes into a positive professor evaluation.

## Getting Started


Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(GGally)
```


### The data

The data were gathered from end of semester student evaluations for a large sample of professors from the University of Texas at Austin. In addition, six students rated the professors' physical appearance. The result is a data frame where each row contains a different course and columns represent variables about the courses and professors. It's called `evals`.

```{r}
glimpse(evals)
```

We have observations on 21 different variables, some categorical and some numerical. The meaning of each variable can be found by bringing up the help file:

```{r help-evals, eval=FALSE}
?evals
```

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.


##        Solution 1:

This is an observational study. Given the study design, it would not be possible to answer the question as it is phrase because its not an experiment and we cant establish the cause. A better rephrased question for this study would be “Is there a relationship between beauty and course evaluations?”


2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?
    
    
##        Solution 2:

```{r}
hist(evals$score, xlab = "Score", main = "Histogram of Score", col = "violet")
```

The distribution of the scores in the Histogram is unimodal left skewed. To a large extent, i expected the students to give the Professor a good evaluation scores given the High reputation of the school. A school with such High reputation will have great Professors. Also, how well the professor performed in lecturing must have also played a role in the evaluation..


3.  Excluding `score`, select two other variables and describe their relationship 
    with each other using an appropriate visualization.
    
    
##        Solution 3:

The mean age of the male student is greater than that of the female student.

```{r}
evals %>% ggplot(aes(x = gender, y = age)) + geom_boxplot(fill = "cyan") + 
  theme_bw() + labs(title = "Barchart of gender vs age of students") + 
  ylab("Age of students in class")
```


## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers are evaluated more favorably. Let's create a scatterplot to see if this appears to be the case:

```{r scatter-score-bty_avg}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point()
```

Before you draw conclusions about the trend, compare the number of observations in the data frame with the approximate number of points on the scatterplot. Is anything awry?

4.  Replot the scatterplot, but this time use `geom_jitter` as your layer. What 
    was misleading about the initial scatterplot?

```{r scatter-score-bty_avg-jitter}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter()
```

##          Solution 4:

**There are overlapped data points in the initial scatterplot which can not be shown as all overlapped points are displayed like a single point.**


5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating. Write out the equation for the linear 
    model and interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?
    
Add the line of the bet fit model to your plot using the following:

##        Solution 5:
    
```{r scatter-score-bty_avg-line-se}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm")
```


<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

1. Equation: y =  3.88034 + 0.06664 * x

2. The slop 0.06664 means for 1 unit increase / decrease of the beauty score, the overall score rating is increased / decreased to change by 0.06664.

3. The `bty_avg` has a p-value 5.08e-05 < 0.05, therefore it is a statistically significant predictor.

4. The correlation coefficient 0.1871424 is very low, which means the relationship between the two variables is weak. Therefore `bty-avg` is not a practically significant predictor.


</div> \hfill\break


```{r}
m_bty <- lm(score~bty_avg, data=evals)
    summary(m_bty)
```
```{r}
cor(evals$score,evals$bty_avg)
```


6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).


##        Solution 6:

```{r}
ggplot(data = m_bty, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")
```


```{r}
ggplot(data = m_bty, aes(x = .resid)) +
  geom_histogram(binwidth = 1) +
  xlab("Residuals")
```

```{r}
ggplot(data = m_bty, aes(sample = .resid)) +
  stat_qq()
```

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

There are four major conditions for the least squares regression.

1. Linearity: It is clearly seen From the scatter plot, that the data follows a linear trend.
2. Nearly Normal residuals: From the histogram, we can see that the residuals are nearly normal
3. Constant Variability: The points residuals vs. fitted plot show that points are somewhat scattered around 0, we could say that there is a constant variability.
4. Independent observations: The data was gathered for a large sample of professors from the University of Texas at Austin and we can safely assume that the observations are independent because the professors are likely different and independent of one another.
</div> \hfill\break


## Multiple linear regression

The data set contains several variables on the beauty score of the professor: individual ratings from each of the six students who were asked to score the physical appearance of the professors and the average of these six scores. Let's take a look at the relationship between one of these scores and the average beauty score.

```{r bty-rel}
ggplot(data = evals, aes(x = bty_f1lower, y = bty_avg)) +
  geom_point()

evals %>% 
  summarise(cor(bty_avg, bty_f1lower))
```

As expected, the relationship is quite strong---after all, the average score is calculated using the individual scores. You can actually look at the relationships between all beauty variables (columns 13 through 19) using the following command:

```{r bty-rels}
evals %>%
  select(contains("bty")) %>%
  ggpairs()
```

These variables are collinear (correlated), and adding more than one of these variables to the model would not add much value to the model. In this application and with these highly-correlated predictors, it is reasonable to use the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score after you've accounted for the professor's gender, you can add the gender term into the model.

```{r scatter-score-bty_avg_pic-color}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.
    
##        solution 7:

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

1. Linearty:
The residuals dispersed most at the upper left of the plot. It doesn't seem to be fully randomly dispersed, but better than the dispersement in question 6.

2. Nearly normal residuals:
The histogram shows a unimodal and left skewed distribution. The distribution of resudials are not normal.

3. Constant variability:
The majority of residuals are distributed between -1 and 1. The constant variability apprears to be met.

Based on the three observation above, the linear model is not reliable.
</div> \hfill\break



```{r}
plot(m_bty_gen$residuals ~ jitter(evals$bty_avg))
abline(h=0)
```


```{r}
boxplot(m_bty_gen$residuals ~ evals$gender)
```


```{r}
hist(m_bty_gen$residuals, breaks=30, col="cyan", xlim=c(-2,1))
```


```{r}
qqnorm(m_bty_gen$residuals)
qqline(m_bty_gen$residuals)
```


8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?
    
##        Solution 8:

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

**Yes, bty_avg still a significant predictor of score. The addition of gender slightly changed the parameter estimate for bty_avg from 5.08e-05 to 6.48e-06, which is still statistically significant for prediction. see summary below**
</div> \hfill\break


```{r}
summary(m_bty)
```

```{r}
summary(m_bty_gen)
```


Note that the estimate for `gender` is now called `gendermale`. You'll see this name change whenever you introduce a categorical variable. The reason is that R recodes `gender` from having the values of `male` and `female` to being an indicator variable called `gendermale` that takes a value of $0$ for female professors and a value of $1$ for male professors. (Such variables are often referred to as "dummy" variables.)

As a result, for female professors, the parameter estimate is multiplied by zero, leaving the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

<!-- We can plot this line and the line corresponding to those with color pictures
with the following  -->
<!-- custom function. -->

```{r twoLines}
ggplot(data = evals, aes(x = bty_avg, y = score, color = pic_color)) +
 geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

9.  What is the equation of the line corresponding to those with color pictures? 
    (*Hint:* For those with color pictures, the parameter estimate is multiplied
    by 1.) For two professors who received the same beauty rating, which color 
    picture tends to have the higher course evaluation score?
    

##        Solution 9:

```{r}
summary (lm(score ~ bty_avg + pic_color, data = evals   ))
```

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

*evaluation score = 4.06318 + 0.05548(bty_avg) - 0.16059 (pic_color)*

*For those with color pictures, the parameter estimate is multiplied by 1 while for those with black and white, the parameter estimate is multiplies by 0. For two professors who received the same beauty rating, balck&white color picture tends to have the higher course evaluation score.*
</div> \hfill\break


The decision to call the indicator variable `gendermale` instead of `genderfemale` has no deeper meaning. R simply codes the category that comes first alphabetically as a $0$. (You can change the reference level of a categorical variable, which is the level that is coded as a 0, using the`relevel()` function. Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.


##        Solution 10:

```{r}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
```

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">
1. For a categorical variable that have N levels, R created N-1 dummy variables for the categorical variable. The level which comes first alphabetically in the categorical variable is treated as a base level by having a coefficient = 0, and no dummy variable is created for it.

2. In this question, two dummy variables `tenure track` and `tenured` are created in the model.  The level `teaching` which comes first alphabetivally in the categorical variable `rank` is treated as a base level.
</div> \hfill\break


The interpretation of the coefficients in multiple regression is slightly different from that of simple regression. The estimate for `bty_avg` reflects how much higher a group of professors is expected to score if they have a beauty rating that is one point higher *while holding all other variables constant*. In this case, that translates into considering only professors of the same rank with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, gender, ethnicity, language of the university where they got their degree, age, proportion of students that filled out evaluations, class size, course level, number of professors, number of credits, average beauty rating, outfit, and picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.
    
##        Solution 11:

*Before runing any statistics, I would expect the number of professors cls_profs to have the highest p-value because I do not expect the number of professors to impact how students rate their professors.*

Let's run the model...

```{r m_full, tidy = FALSE}
m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.
    
##        Solution 12:

**My suspicions are correct, cls_profs has the highest p_value.**


13. Interpret the coefficient associated with the ethnicity variable.


##        Solution 13:

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

**The value of the coefficient `ethnicity not minority` is 0.1234929 means a professers who re not ethnicity minorities have overall score 0.1234929 higher than those who are ethnicity minorities, keeping all other variable constant.**
</div> \hfill\break


14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?
    
    
##        Solution 14:

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

**The coefficients and significances of other variables are slightly changed. This means the collinearty of the dropped variable to the other variables is not significant.**
</div> \hfill\break

```{r}
m_drop_1 <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_drop_1)
```


15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.


##        solution 15:

```{r}
m_step_final <- lm(formula = score ~ ethnicity + gender + language + age + 
                    cls_perc_eval + cls_credits + bty_avg + pic_color, data = evals)
summary(m_step_final)
```
**The linear model for predicting the score based on the final model that I settled on is given by:**

eval_score = 3.772 + 0.207(gender) + 0.168(ethnicity) - 0.206(language) - 00.6(age) + 0.005(cls_perc_eval) + 0.505(cls_credits) + 0.051(bty_avg) - 0.191(pic_color)


16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.
    
    
##        Solution 16:

```{r}
par(mfrow=c(1,2))
plot(m_step_final,c(1,2))
```


```{r}
par(mfrow=c(1,3))
plot(jitter(evals$age), resid(m_step_final))
abline(h=0, col="violet")
plot(jitter(evals$cls_perc_eval), resid(m_step_final))
abline(h=0, col="violet")
plot(jitter(evals$bty_avg), resid(m_step_final))
abline(h=0, col="violet")
```


```{r}
hist(resid(m_step_final))
```

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

1. Linearty:
For the quantitative variables `age`, `cls_perc_eval`, `bty_avg`: The residuals are most likely to be randomly dispersed, no obvious shapes or patterns are found.

2. Nearly normal residuals
The histogram of the residuals shows a unimodal and left skewed distribution. The qq plot shows the residuals are mostly line along on the normal line.The normal residual condiction is somewhat met.

3. Constant variability
The majority of residuals are distributed between -1 and 1. The constant variability apprears to be met.

Based on the three observation above, the linear model is reliable.
</div> \hfill\break


17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?
    
    
##        Solution 17:

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

**This condition will break the assumption that all sample cases are randomly collected and are independent of each other.**
</div> \hfill\break


18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.
    
##        Solution 18:

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

The final model contains 8 variables including `ethnicity`, `gender`, `language`, `age`,`cls_prec_eval`,`cls_credits`,`bty_avg`,`pic_color`. 

*According to the model, the characteristics of a professor and course at University of Texas at Austin that would be associated with a high evaluation score would be a young male Professor who is not minority, speaks English, with a black and white profile picture, and considered handsome.*
</div> \hfill\break


19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?
    
    
##        Solution 19:

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

*The data is collected from single university which can not represent all universities. Other conditions that may be unique to other universities where not considered. Therefore the model that is built on this data can not be applied to other universities.*

</div> \hfill\break




* * *
